**Evaluation of machine learning methods for 16S rRNA gene data**

### 2019-06-11 submission/manuscript.Rmd
changes on train time

######################################################################
#-------------------------- Load walltime data -----------------#
######################################################################
source("../code/learning/load_walltime.R")

######################################################################
#---------------------- Load AUROC performance data -----------------#
######################################################################
source("../code/learning/load_AUROC_data.R")
```
\vspace{35mm}

@@

Machine learning (ML) modeling of the human microbiome has the potential to identify the microbial biomarkers and aid in diagnosis of many chronic diseases such as inflammatory bowel disease, diabetes and colorectal cancer. Progress has been made towards developing ML models that predict health outcomes from bacterial abundances, but rigourous ML models are scarce in part due to the flawed modeling methods that call into question the validity of developed ML models. Furthermore, using black box ML models has hindered the validation of microbial biomarkers. To overcome these challenges, we benchmarked seven different ML models that use fecal 16S rRNA sequences to predict the presence/absence of colorectal cancer (CRC) lesions (n=490 patients, 261 controls and 229 cases). To show the effect of model selection, we assessed the accuracy, interpretability, and computational efficiency of the following models: L2-regularized logistic regression, L1 and L2 support vector machines (SVM) with linear and radial basis function kernels, a decision tree, random forest, and extreme gradient boosting (XGBoost). The random forest model was best at detecting CRC lesions with an AUROC of 0.695 but it was slow to train (101.3 h) and hard to interpret. Despite its simplicity, L2-regularized logistic regression followed random forest in predictive performance with an AUROC of 0.680, and it trained the fastest (12 min). This study showed that we should choose ML models based on our expectations of performance and interpretability as well as our  computational resources. It also established standards for modeling pipelines of microbiome-associated ML models.

@@

To showcase a reliable ML pipeline and to shed light on how much ML model selection can affect modeling results, we performed an empirical analysis comparing several different ML models using the same dataset and with a robust ML pipeline. We used a previously published colorectal cancer (CRC) study [@baxter_dna_2016] which had fecal 16S rRNA gene sequences from `r paste(nrow(shared))` patients. CRC is the third most common form of cancer worldwide in which the human-associated human microbiome is hypothesized to directly contribute to its development. We built ML models using fecal 16S rRNA gene sequences to predict patients with normal colons or patients with colonic tumors which are called screen relevant neoplasias (SRN). The study had `r paste(sum(sum(meta$dx=="normal")+sum(meta$Dx_Bin=="Adenoma")))` normal and `r paste(sum(sum(meta$dx=="cancer")+sum(meta$Dx_Bin=="adv Adenoma")))` SRN samples. We established modeling pipelines for L2-regularized logistic regression, L1 and L2 support vector machines (SVM) with linear and radial basis function kernels, a decision tree, random forest and XGBoost. Our ML pipeline utilized held-out test data to evaluate generalization performance of each ML model. The median test AUROC varied from `r paste(round(test_all[performance_index[1],3], 3))`  to `r paste(round(test_all[performance_index[7],3], 3))`. Random forest had the highest median AUROC for detecting SRN. Despite its simplicity, the L2-regularized logistic regression followed random forest in performance. In terms of computational efficiency, L2 logistic regression trained the fastest (`r paste(round(walltime_df[walltime_index[1],2], 3))` hours, std ± `r paste(round(walltime_df[walltime_index[1],3], 3))`), while XGBoost took the longest (`r paste(round(walltime_df[walltime_index[7],2], 3))` hours, std ± `r paste(round(walltime_df[walltime_index[7],3], 3))`). We found that median cross-validation and testing AUROC varied `r paste(round(abs(difference[difference_index[7]]), 3))`, which highlights the importance of a seperate held-out test set and consistent preprocessing of the data prior to evaluation. Aside from evaluating generalization and classification performances for each of these models, this study established standards for modeling pipelines of microbiome-associated machine learning models.

@@

Linear models trained faster than non-linear models. L2 logistic Rregression and L1 and L2 SVM with linear kernel had training times of `r paste(round(walltime_df[walltime_index[1],2], 1))` hours, (std ± `r paste(round(walltime_df[walltime_index[1],3], 2))`), `r paste(round(walltime_df[walltime_index[2],2], 1))` hours, (std ± `r paste(round(walltime_df[walltime_index[2],3],2))`), and `r paste(round(walltime_df[walltime_index[3],2], 1))` hours, (std ± `r paste(round(walltime_df[walltime_index[3],3], 2))`), respectively. Whereas, SVM with radial basis function kernel, decision tree, random forest and xgboost had training times of `r paste(round(walltime_df[walltime_index[4],2], 1))` hours, (std ± `r paste(round(walltime_df[walltime_index[4],3], 1))`), `r paste(round(walltime_df[walltime_index[5],2], 1))` hours, (std ± `r paste(round(walltime_df[walltime_index[5],3], 1))`), `r paste(round(walltime_df[walltime_index[6],2], 1))` hours, (std ± `r paste(round(walltime_df[walltime_index[6],3], 1))`) and `r paste(round(walltime_df[walltime_index[7],2], 1))` hours, (std ± `r paste(round(walltime_df[walltime_index[7],3], 1))`), respectively [Figure 4].

@@

\newpage
\includegraphics{Figure_1}
**Figure 1. Machine learning pipeline showing predictive model training and evaluation flowchart.  **  We split the data 80%/20% stratified to maintain the overall label distribution, performed five-fold cross-validation on the training data to select the best hyperparameter setting and then using these hyperparameters to train all of the training data. The model was evaluated on a held-out set of data (not used in selecting the model).
Abbreviations: AUROC, area under the receiver operating characteristic curve
\newpage
\includegraphics{Figure_2.png}
**Figure 2. Generalization and classification performance of ML models using AUROC values of all cross validation and testing performances. **   The median AUROC for diagnosing individuals with SRN using bacterial abundances was higher than chance (depicted by horizontal line at 0.50) for all the ML models. Discriminative performance of random forest model was higher than other ML models. The boxplot shows quartiles at the box ends and the statistical median as the horizontal line in the box. The whiskers show the farthest points that are not outliers. Outliers are data points that are not within 3/2 times the interquartile ranges. 
Abbreviations: SRN, screen-relevant neoplasias; AUROC, area under the receiver operating characteristic curve; SVM, support vector machine; XGBoost, extreme gradient boosting
\newpage
\includegraphics{Figure_3.png}
**Figure 3. Interpretation of the linear ML models.** (A) L2 logistic regression coefficients  (B) L1 SVM with linear kernel feature weights (C) L2 SVM with linear kernel feature weights. The means weights and coefficients of the most important five OTUs for each model are shown here with the standard deviation over 100 data-splits. Similar OTUs had the largest impact on the predictive performance of L2 logistic regression and L2 SVM with linear kernel. 
Abbreviations: SVM, support vector machine; OTU, Operational Taxonomic Unit.
\newpage
\includegraphics{Figure_4.png}
**Figure 4. Explanation of the non-linear ML models.** (A) SVM with radial basis kernel (B) decision tree (C) random forest (D) XGboost feaure importances were explanied using permutation importance using held-out test set. The gray rectangle and the dashed line show the IQR range and median of the base testing AUROC without any permutation performed. For all the tree-based models, a *Peptostreptococcus* species (OTU00367) had the largest impact on predictive performance of the model. 
Abbreviations: SVM, support vector machine; OTU, Operational Taxonomic Unit; RBF, radial basis kernel; OTU, Operational Taxonomic Unit.
\newpage
\includegraphics{Figure_5.png}
**Figure 5. Computational efficiency of seven ML models.** The walltimes for training and testing of each data-split showed the differences in computational efficieny of the seven models. The median walltime in hours was the highest for XGBoost and shortest for L2 logistic regression. The boxplot shows quartiles at the box ends and the statistical median as the horizontal line in the box. The whiskers show the farthest points that are not outliers. Outliers are data points that are not within 3/2 times the interquartile ranges. 
Abbreviations: AUROC, area under the receiver operating characteristic curve; SVM, support vector machine; XGBoost, extreme gradient boosting.  
\newpage
\includegraphics{Figure_S1.png}
**Figure S1. Hyperparameter setting performances for linear models.** 
Abbreviations: 
\newpage
\includegraphics{Figure_S2.png}
**Figure S2. Hyperparameter setting performances for non-linear models.** 
Abbreviations: 
\newpage

>>>>

######################################################################
#-------------------------- Load traintime data -----------------#
######################################################################
source("../code/learning/load_traintime.R")

######################################################################
#-------- Load AUROC performance data and Wilcoxon test ------------#

######################################################################
source("../code/learning/load_AUROC_data.R")
```
\vspace{35mm}

@@

Machine learning (ML) modeling of the human microbiome has the potential to identify the microbial biomarkers and aid in diagnosis of many chronic diseases such as inflammatory bowel disease, diabetes and colorectal cancer. Progress has been made towards developing ML models that predict health outcomes from bacterial abundances, but rigourous ML models are scarce in part due to the flawed modeling methods that call into question the validity of developed ML models. Furthermore, using black box ML models has hindered the validation of microbial biomarkers. To overcome these challenges, we benchmarked seven different ML models that use fecal 16S rRNA sequences to predict the presence/absence of colorectal cancer (CRC) lesions (n=490 patients, 261 controls and 229 cases). To show the effect of model selection, we assessed the accuracy, interpretability, and computational efficiency of the following models: L2-regularized logistic regression, L1 and L2 support vector machines (SVM) with linear and radial basis function kernels, a decision tree, random forest, and extreme gradient boosting (XGBoost). The random forest model was best at detecting CRC lesions with an AUROC of 0.695 but it was slow to train (101.3 h) and hard to interpret. Despite its simplicity, L2-regularized logistic regression followed random forest in predictive performance with an AUROC of 0.680, and it trained much faster (12 min). This study showed that we should choose ML models based on our expectations of performance and interpretability as well as our  computational resources. It also established standards for modeling pipelines of microbiome-associated ML models.

@@

To showcase a reliable ML pipeline and to shed light on how much ML model selection can affect modeling results, we performed an empirical analysis comparing several different ML models using the same dataset and with a robust ML pipeline. We used a previously published colorectal cancer (CRC) study [@baxter_dna_2016] which had fecal 16S rRNA gene sequences from `r paste(nrow(shared))` patients. CRC is the third most common form of cancer worldwide in which the human-associated human microbiome is hypothesized to directly contribute to its development. We built ML models using fecal 16S rRNA gene sequences to predict patients with normal colons or patients with colonic tumors which are called screen relevant neoplasias (SRN). The study had `r paste(sum(sum(meta$dx=="normal")+sum(meta$Dx_Bin=="Adenoma")))` normal and `r paste(sum(sum(meta$dx=="cancer")+sum(meta$Dx_Bin=="adv Adenoma")))` SRN samples. We established modeling pipelines for L2-regularized logistic regression, L1 and L2 support vector machines (SVM) with linear and radial basis function kernels, a decision tree, random forest and XGBoost. Our ML pipeline utilized held-out test data to evaluate generalization performance of each ML model. The median test AUROC varied from `r paste(round(test_all[performance_index[1],3], 3))`  to `r paste(round(test_all[performance_index[7],3], 3))`. Random forest had the highest median AUROC for detecting SRN. Despite its simplicity, the L2-regularized logistic regression followed random forest in performance. In terms of computational efficiency, L1 SVM with linear kernel trained the fastest (`r paste(round(traintime_df[traintime_index[1],2], 3))` hours, std ± `r paste(round(traintime_df[traintime_index[1],3], 3))`), while XGBoost took the longest (`r paste(round(traintime_df[traintime_index[7],2], 3))` hours, std ± `r paste(round(traintime_df[traintime_index[7],3], 3))`). We found that median cross-validation and testing AUROC varied `r paste(round(abs(difference[difference_index[7]]), 3))`, which highlights the importance of a seperate held-out test set and consistent preprocessing of the data prior to evaluation. Aside from evaluating generalization and classification performances for each of these models, this study established standards for modeling pipelines of microbiome-associated machine learning models.

@@

Linear models trained faster than non-linear models. L2 logistic Rregression and L1 and L2 SVM with linear kernel had training times of `r paste(round(traintime_df[traintime_index[1],2], 1))` hours, (std ± `r paste(round(traintime_df[traintime_index[1],3], 2))`), `r paste(round(traintime_df[traintime_index[2],2], 1))` hours, (std ± `r paste(round(traintime_df[traintime_index[2],3],2))`), and `r paste(round(traintime_df[traintime_index[3],2], 1))` hours, (std ± `r paste(round(traintime_df[traintime_index[3],3], 2))`), respectively. Whereas, SVM with radial basis function kernel, decision tree, random forest and xgboost had training times of `r paste(round(traintime_df[traintime_index[4],2], 1))` hours, (std ± `r paste(round(traintime_df[traintime_index[4],3], 1))`), `r paste(round(traintime_df[traintime_index[5],2], 1))` hours, (std ± `r paste(round(traintime_df[traintime_index[5],3], 1))`), `r paste(round(traintime_df[traintime_index[6],2], 1))` hours, (std ± `r paste(round(traintime_df[traintime_index[6],3], 1))`) and `r paste(round(traintime_df[traintime_index[7],2], 1))` hours, (std ± `r paste(round(traintime_df[traintime_index[7],3], 1))`), respectively [Figure 4].

@@

\newpage
\includegraphics{Figure_1}
**Figure 1. Machine learning pipeline showing predictive model training and evaluation flowchart.  **  We split the data 80%/20% stratified to maintain the overall label distribution, performed five-fold cross-validation on the training data to select the best hyperparameter setting and then using these hyperparameters to train all of the training data. The model was evaluated on a held-out set of data (not used in selecting the model).
Abbreviations: AUROC, area under the receiver operating characteristic curve
\newpage
\includegraphics{Figure_2.png}
**Figure 2. Generalization and classification performance of ML models using AUROC values of all cross validation and testing performances. **   The median AUROC for diagnosing individuals with SRN using bacterial abundances was higher than chance (depicted by horizontal line at 0.50) for all the ML models. Discriminative performance of random forest model was higher than other ML models. The boxplot shows quartiles at the box ends and the statistical median as the horizontal line in the box. The whiskers show the farthest points that are not outliers. Outliers are data points that are not within 3/2 times the interquartile ranges. 
Abbreviations: SRN, screen-relevant neoplasias; AUROC, area under the receiver operating characteristic curve; SVM, support vector machine; XGBoost, extreme gradient boosting
\newpage
\includegraphics{Figure_3.png}
**Figure 3. Interpretation of the linear ML models.** (A) L2 logistic regression coefficients  (B) L1 SVM with linear kernel feature weights (C) L2 SVM with linear kernel feature weights. The means weights and coefficients of the most important five OTUs for each model are shown here with the standard deviation over 100 data-splits. Similar OTUs had the largest impact on the predictive performance of L2 logistic regression and L2 SVM with linear kernel. 
Abbreviations: SVM, support vector machine; OTU, Operational Taxonomic Unit.
\newpage
\includegraphics{Figure_4.png}
**Figure 4. Explanation of the non-linear ML models.** (A) SVM with radial basis kernel (B) decision tree (C) random forest (D) XGboost feaure importances were explanied using permutation importance using held-out test set. The gray rectangle and the dashed line show the IQR range and median of the base testing AUROC without any permutation performed. For all the tree-based models, a *Peptostreptococcus* species (OTU00367) had the largest impact on predictive performance of the model. 
Abbreviations: SVM, support vector machine; OTU, Operational Taxonomic Unit; RBF, radial basis kernel; OTU, Operational Taxonomic Unit.
\newpage
\includegraphics{Figure_5.png}
**Figure 5. Computational efficiency of seven ML models.** The traintimes for training and testing of each data-split showed the differences in computational efficieny of the seven models. The median traintime in hours was the highest for XGBoost and shortest for L2 logistic regression. The boxplot shows quartiles at the box ends and the statistical median as the horizontal line in the box. The whiskers show the farthest points that are not outliers. Outliers are data points that are not within 3/2 times the interquartile ranges. 
Abbreviations: AUROC, area under the receiver operating characteristic curve; SVM, support vector machine; XGBoost, extreme gradient boosting.  
\newpage
\includegraphics{Figure_S1.png}
**Figure S1. Hyperparameter setting performances for linear models.** 
Abbreviations: 
\newpage
\includegraphics{Figure_S2.png}
**Figure S2. Hyperparameter setting performances for non-linear models.** 
Abbreviations: 
\newpage
